Global Life Expectancy: Unraveling Health and Economic Determinants (2015-2020)

### 2024-12-04 paper/paper.qmd
ini
jk

---
title: "Global Life Expectancy: Unraveling Health and Economic Determinants (2015-2020)"
subtitle: "Multiple Linear Regression Analyzing Critical Factors Shaping Life Expectancy "
author: 
  - Yanfei Huang
thanks: "Code and data are available at: [https://github.com/wendyhuan/lifeexpectancy](https://github.com/wendyhuan/lifeexpectancy)."
date: 02 December 2024
date-format: long
abstract: "This paper analyze life expectancy and its determinants through 184 countries and make the prediction of people from different Income Group. Multiple Linear regression is used to deploying the life expectanc with gender, income and region. Predictions of life expectancy of people from different income group is made according to these essential predictors. The finding indicates that the life expectancy is tend to get higher as the economic class grows. And we predict that the average age of person from low, lower-middle, upper-middle and high are 65,70, 75,80. The result of this paper will help in suggesting a country which area should be given importance in order to efficiently improve the life expectancy of its population."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

# load data
lelecleaned_data <- read_parquet("~/lifeexpectancy/data/02-analysis_data/lelecleaned.parquet")
lea60 <- read_parquet("~/lifeexpectancy/data/02-analysis_data/lea60.parquet")
leab <- read_parquet("~/lifeexpectancy/data/02-analysis_data/leab.parquet")
male_table <- read_parquet("~/lifeexpectancy/data/02-analysis_data/male_table.parquet")
female_table <- read_parquet("~/lifeexpectancy/data/02-analysis_data/female_table.parquet")
average_life_expectancy <- read_parquet("~/lifeexpectancy/data/02-analysis_data/average_life_expectancy.parquet")
sum_sta <- read_parquet("~/lifeexpectancy/data/02-analysis_data/sum_sta.parquet")
matable <- read_parquet("~/lifeexpectancy/data/02-analysis_data/matable.parquet")
femaletable <- read_parquet("~/lifeexpectancy/data/02-analysis_data/femaletable.parquet")
model_developed <- readRDS(file = "~/lifeexpectancy/models/ledata_normal_model.rds")
```

@@

This analysis uses the statistical programming language R [@citeR] and several libraries, including `tidyverse` [@tidyverse], `janitor` [@janitor], `knitr` [@knitr], `dplyr` [@dplyr], `arrow` [@arrow], `purrr` [@purrr], `sf` [@sf], and `here` [@here] for data manipulation. `ggplot2` [@ggplot], `ggcorrplot` [@ggcorrplot] and `kableExtra` [@kableExtra] for visualization. The dataset covers various predictors conducted across multiple countries, capturing the support for a country to determine the predicting factor which is contributing to lower value of life expectancy.

@@

**Region(Region)**: This is a categorical variable that classifies the geographic location of the countries. It is mapped one by one by the name of the country. The data is stored as the name of the continent(‘Africa’, ‘Oceania’, ‘Asia’, ‘Europe’, ‘North America’, ‘South America’).

@@

@fig-sum_sta shows the average of life expectancy under male, female, 6 different regions and 4 income groups. Life expectancy according to different countries have attached and be found in data cleaning full data in @sec-appendix.

@@

```{r}
#| label: tbl-sum_sta
#| fig-cap: "This table shows the cleaned data of life expectancy at different predictor. We could visual that high income has the highest life expectancy of 79.8, compared to lower income has the least life expectancy of 62.35. Among all the region, it's clear that Africa has the least 64.17, while Europe has the highest 78.50 years old. Female has slightly higher life expectancy than male. "
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Format and display the sum_sta table
# Load necessary libraries
library(knitr)
library(kableExtra)

# Format and style the sum_sta table using kableExtra
sum_sta %>%
  kbl(
    col.names = c("Region/Income/Gender", "Life Expectancy"),  # Column names
    align = c("l", "c"),                                       # Left-align column 1, center-align column 2
    digits = 2,                                                # Round values to 2 decimal places
    booktabs = TRUE,                                           # LaTeX-style table rule
  ) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),             # Add striped rows and prevent table floating
    full_width = FALSE,                                        # Prevent table from stretching to full width
    font_size = 10                                             # Adjust font size
  ) %>%
  add_header_above(c(" " = 1, "Life Expectancy Summary" = 1),bold = TRUE) %>%
  add_header_above(c(" " = 2), bold = TRUE) # Add a header row

```
## Outcome Variables

@@

1.  **Income Group(`Income_Group`)**: The income group of a country is the key factor influencing the life expectancy. This variable provides context for comparing life expectancy between different levels of country income group. The division of the life expectancy by the income group of a country because income levels often correlate strongly with various factors affecting health and longevity. Higher income countries typically have more resources to invest in robust healthcare systems, with better living conditions while lower income countries often face challenges such as inadequate healthcare infrastructure and malnutrition.

Following by the four income group, we would like to see the distribution of the income group of 184 WHO member countries. @fid-income shows that there are 55 countries at the high income group, 26 countries at low income group, 54 countries at the lower-middle group while there are 49 countries at the upper-middle group. With significant less low income countries, we expect to see the life expectancy lie in a comparably higher range. On the other hand, there might exist sever outliers dropping the mean.

@@

```{r}
#| label: fig-income
#| fig-cap: "Distribution of country income group."
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| label: fig-summary
#| fig-cap: "Summary statistics of the number of life expectancy over countries and years. "
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Calculate summary statistics
life_expectancy_stats <- leab %>%
  summarise(
    Mean = round(mean(`Life Expectancy`, na.rm = TRUE),1),
    Median = median(`Life Expectancy`, na.rm = TRUE),
    Max = max(`Life Expectancy`, na.rm = TRUE),
    SD = round(sd(`Life Expectancy`, na.rm = TRUE),1),
    Variance = round(var(`Life Expectancy`, na.rm = TRUE),1),
    N = sum(!is.na(`Life Expectancy`)) # Count of non-missing values
  )

#Create and format the table using kableExtra
life_expectancy_stats %>%
  kbl(
    caption = "Summary Statistics of Life Expectancy",
    col.names = c("Mean", "Median", "Max", "Standard Deviation", "Variance", "N"),
    align = "c",
    digits = 2 # Ensure numeric values are rounded to 2 decimal places
  ) %>%
  kable_styling(
    full_width = FALSE, 
    bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
    font_size = 14
  ) %>%
  column_spec(1, bold = TRUE, color = "white", background = "darkblue") %>% # Highlight the first column
  column_spec(2:ncol(life_expectancy_stats), color = "black") %>% # Style the data columns
  row_spec(0, bold = TRUE, background = "lightgray") %>% # Add a header row styling
  add_header_above(c(" " = 6)) %>% # Add optional grouping header
  footnote(
    general = "This table summarizes the key statistics of life expectancy data.",
    footnote_as_chunk = TRUE, 
    general_title = "Note:"
  )

@@

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \text{(1)} \\
\mu_i &= \alpha + \beta_1 \text{Region}_i + \beta_2 \text{Gender}_i + \beta_3 \text{Income_Group}_i \\text{(2)} \\
\alpha &\sim \text{Normal}(0, 2.5) \text{(3)}\\
\beta_1, \beta_2, \beta_3 &\sim \text{Normal}(0, 2.5) \text{(4)}\\
\sigma &\sim \text{Exponential}(1) \text{(5)} \\
\end{align}
```

Where:

@@

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`. More explanation of the model could be found [Appendix -@sec-model-details].

### Model justification

@@

#### Model Validation

@@

```{r}
# Split the dataset into training and testing sets (80% train, 20% test)
set.seed(123)  # Setting seed for reproducibility
train_index <- createDataPartition(leab$`Life Expectancy`, p = 0.8, list = FALSE)

@@

# Predict life expectancy using the test data
predictions <- predict(model_developed, newdata = test_data)
```

```{r}
#| label: tbl-modelpredict
#| tbl-cap: Summary of the life expectancy model, which includes gender, region and income group. The table presents the model RSME and out-of sample testing.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# Create a styled table with kableExtra
rmse_table %>%
  kbl(
    col.names = c("Dataset", "RMSE"), 
    caption = "RMSE Comparison for Test and Training Data",
    align = "c", # Center alignment for columns
    digits = 2   # Display RMSE values with two decimal places
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  column_spec(1, bold = TRUE, color = "white", background = "steelblue") %>%
  column_spec(2, color = "black") %>%
  add_header_above(c(" " = 1, "Root Mean Squared Error (RMSE)" = 1))

@@

# Results

@@

# Create a model summary with median absolute deviation (mad) for statistics
modelsummary(
  list("Gaussian(Normal)" = model_developed),
  statistic = "mad",  # Median absolute deviation
  fmt = 2             # Format results to 2 decimal places
)

@@

With an R-squared value of 0.976, the model demonstrates a strong level of explanatory power, supporting the accuracy of the results. The adjusted R-squared value is 0.975, which is nearly identical to the R-squared value, indicating the model's robustness with 3,312 observations. The RMSE of 4.37 reflects the average magnitude of the model's prediction errors, suggesting good predictive accuracy. Although values like LOOIC and WAIC are excluded due to the lack of alternative models for comparison, the overall results indicate that the model performs well on the cleaned dataset.Further evaluation of the model results will be done in @sec-model_details.

# Discussion

@@

From the summary and analysis of @tbl-regresult, we have seen that high income group has the significant high life expectancy compared to the lower income group, which indicates that life expectancy is significantly influenced by income groups. It is known that income determines access to resources, healthcare and living conditions. Difference in life expectancy across income groups reflect broader inequalities in socioeconomic factors and highlight disparities in global health outcomes.

@@

# Appendix {#sec-appendix}

@@

## Data Cleaning Full Table

```{r}
#| label: tbl-cleaned_average
#| tbl-cap: "Full cleaned Average life expectancy"
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create a nicely formatted kable table
average_life_expectancy %>%
  kable(
    col.names = c("Country", "Gender", "Income Group", "Region", "Average Life Expectancy"),
    align = c("l", "l", "l", "l", "r"),
    digits = 2
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#0073C2") %>% # Style the header row
  column_spec(5, bold = TRUE, background = "#F7F7F7")

@@

**1. Outcome variable (yi)**

@@

**2. Linear Predictor (μi)**

The linear predictor $μ_i$ is a linear combination of the predictors, which includes an intercept term $α$ and the coefficients $β_1$, $β_2$, $β_3$, corresponding to Region, Income Group, and Gender, respectively. This linear equation models the relationship between life expectancy and the predictors $\mu_i = \alpha + \beta_1(Region) + \beta_2(Income Group) + \beta_3 (Gender)$ The values for these coefficients (betas) represent the expected change in life expectancy for a one-unit change in each respective predictor.

@@

**4. Residual Standard Deviation (σ)** The model also estimates $σ$, which represents the standard deviation of the residuals — the errors between the observed life expectancy values and those predicted by the model. The prior for $σ$ is assumed to follow an exponential distribution with a rate of 1, meaning that smaller values of $σ$ (indicating less variation) are somewhat more likely than larger ones.

@@

```{r}
#| label: tbl-testpredicted
#| tbl-cap: "Life expectancy over countries and years"
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# Format the table using kable
kable(
  results_df,
  col.names = c("Country", "Actual Life Expectancy", "Predicted Life Expectancy", "RMSE (Test)", "RMSE (Train)", "Indication"),
  align = c("l", "r", "r", "r", "r", "l"),
  digits = 2,
  booktabs = TRUE
)

>>>>

---
title: "Global Life Expectancy: Unraveling Health and Economic Determinants (2015-2020)"
subtitle: "Multiple Linear Regression Analyzing Critical Factors Shaping Life Expectancy "
author: 
  - Yanfei Huang
thanks: "Code and data are available at: [https://github.com/wendyhuan/lifeexpectancy](https://github.com/wendyhuan/lifeexpectancy)."
date: 02 December 2024
date-format: long
abstract: "This paper analyze life expectancy and its determinants through 184 countries and make the prediction of people from different Income Group. Multiple Linear regression is used to deploying the life expectanc with gender, income and region. Predictions of life expectancy of people from different income group is made according to these essential predictors. The finding indicates that the life expectancy is tend to get higher as the economic class grows. And we predict that the average age of person from low, lower-middle, upper-middle and high are 62.35, 68.40, 73.38, 79.38. The result of this paper will help in suggesting a country which area should be given importance in order to efficiently improve the life expectancy of its population."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

# load data
lelecleaned_data <- read_parquet("~/lifeexpectancy/data/02-analysis_data/lelecleaned.parquet")
lea60 <- read_parquet("~/lifeexpectancy/data/02-analysis_data/lea60.parquet")
leab <- read_parquet("~/lifeexpectancy/data/02-analysis_data/leab.parquet")
male_table <- read_parquet("~/lifeexpectancy/data/02-analysis_data/male_table.parquet")
female_table <- read_parquet("~/lifeexpectancy/data/02-analysis_data/female_table.parquet")
average_life_expectancy <- read_parquet("~/lifeexpectancy/data/02-analysis_data/average_life_expectancy.parquet")
sum_sta <- read_parquet("~/lifeexpectancy/data/02-analysis_data/sum_sta.parquet")
matable <- read_parquet("~/lifeexpectancy/data/02-analysis_data/matable.parquet")
femaletable <- read_parquet("~/lifeexpectancy/data/02-analysis_data/femaletable.parquet")
model_developed <- readRDS(file = "~/lifeexpectancy/models/ledata_normal_model.rds")
bothg_average_life_expectancy <- read_parquet("~/lifeexpectancy/data/02-analysis_data/bothg_average_life_expectancy.parquet")
```

@@

This analysis uses the statistical programming language R [@citeR] and several libraries, including `tidyverse` [@tidyverse], `janitor` [@janitor], `knitr` [@knitr], `dplyr` [@dplyr], `arrow` [@arrow], `purrr` [@purrr], `sf` [@sf], and `here` [@here] for data manipulation. `ggplot2` [@ggplot2], `ggcorrplot` [@ggcorrplot] and `kableExtra` [@kableExtra] for visualization. The dataset covers various predictors conducted across multiple countries, capturing the support for a country to determine the predicting factor which is contributing to lower value of life expectancy.

@@

**Region(`Region`)**: This is a categorical variable that classifies the geographic location of the countries. It is mapped one by one by the name of the country. The data is stored as the name of the continent(‘Africa’, ‘Oceania’, ‘Asia’, ‘Europe’, ‘North America’, ‘South America’).

@@

@tbl-sum_sta shows the average of life expectancy under male, female, 6 different regions and 4 income groups. Life expectancy according to different countries have attached and be found in data cleaning full data in @sec-appx.

@@

```{r}
#| label: tbl-sum_sta
#| tbl-cap: "This table shows the cleaned data of life expectancy at different predictor. We could visual that high income has the highest life expectancy of 79.8, compared to lower income has the least life expectancy of 62.35. Among all the region, it's clear that Africa has the least 64.17, while Europe has the highest 78.50 years old. Female has slightly higher life expectancy than male. "
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Format and style the sum_sta table using kableExtra
sum_sta %>%
  kable(
    col.names = c("Region/Income/Gender", "Life Expectancy"),  # Column names
    align = c("l", "c"),                                       # Left-align column 1, center-align column 2
    digits = 2 ) %>%
  kable_styling(          # Add striped rows and prevent table floating
    full_width = FALSE,                                        # Prevent table from stretching to full width
    font_size = 10                                             # Adjust font size
  )

## Outcome Variables

@@

1.  **Income Group(`Income_Group`)**: The income group of a country is the key factor influencing the life expectancy. This variable provides context for comparing life expectancy between different levels of country income group. The division of the life expectancy by the income group of a country because income levels often correlate strongly with various factors affecting health and longevity. Higher income countries typically have more resources to invest in robust healthcare systems, with better living conditions while lower income countries often face challenges such as inadequate healthcare infrastructure and malnutrition. Following by the four income group, we would like to see the distribution of the income group of 184 WHO member countries. @fig-income shows that there are 55 countries at the high income group, 26 countries at low income group, 54 countries at the lower-middle group while there are 49 countries at the upper-middle group. With significant less low income countries, we expect to see the life expectancy lie in a comparably higher range. On the other hand, there might exist sever outliers dropping the mean.

@@

```{r}
#| label: fig-income
#| fig-cap: "Distribution of country income group. Among 184 WHO member countries, there are 55 high income countries, 54 lower middle income countries and 49 upper middle income countries. There are significant less lower income countries, which we could assume there might exist less outliers in life experctanncy."
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| label: fig-summary
#| fig-cap: "Summary statistics of the number of life expectancy over countries and years. Less difference between mean and median, showing that the life expectancy distribution might shows a naturally bell curve.However, variance is slightly high indicates that the tails on both sides might be long. "
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Calculate summary statistics
life_expectancy_stats <- leab %>%
  summarise(
    Mean = round(mean(`Life Expectancy`, na.rm = TRUE), 1),
    Median = median(`Life Expectancy`, na.rm = TRUE),
    Max = max(`Life Expectancy`, na.rm = TRUE),
    SD = round(sd(`Life Expectancy`, na.rm = TRUE), 1),
    Variance = round(var(`Life Expectancy`, na.rm = TRUE), 1),
    N = sum(!is.na(`Life Expectancy`)) # Count of non-missing values
  )

#Create and format the table using kableExtra
life_expectancy_stats %>%
  kable(
    col.names = c("Mean", "Median", "Max", "Standard Deviation", "Variance", "N"),
    align = "c",
    digits = 2 # Ensure numeric values are rounded to 2 decimal places
  ) %>%
  kable_styling(
    full_width = FALSE, 
    font_size = 14
  )

@@

```{=tex}

\begin{align*} 
y_i|\mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_1 \text{Region}_i + \beta_2 \text{Gender}_i + \beta_3 \text{Income\_Group}_i \\
\alpha &\sim \text{Normal}(0, 2.5) \\
\beta_1, \beta_2, \beta_3 &\sim \text{Normal}(0, 2.5) \\
\sigma &\sim \text{Exponential}
\end{align*}

```
Where:

@@

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`. More explanation of the model could be found [Appendix @sec-model-details].

## Model justification {#sec-modjust}

@@

## Model Validation

@@

```{r}
#| label: tbl-modelpredict
#| tbl-cap: "Summary of the life expectancy model, which includes gender, region and income group. The table presents the model RSME and out-of sample testing. Seeing the testing data slightly less than the training data, our model is under good generalization."
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Split the dataset into training and testing sets (80% train, 20% test)
set.seed(123)  # Setting seed for reproducibility
train_index <- createDataPartition(leab$`Life Expectancy`, p = 0.8, list = FALSE)

@@

# Predict life expectancy using the test data
predictions <- predict(model_developed, newdata = test_data)

@@

# Create a styled table with kableExtra
rmse_table %>%
  kable(
    col.names = c("Dataset", "RMSE"), 
    align = "c", # Center alignment for columns
    digits = 2   # Display RMSE values with two decimal places
  ) %>%
  kable_styling(
    full_width = FALSE,
    font_size = 14
  )

@@

# Results {#sec-result}

@@

# Create a model summary with median absolute deviation (mad) for statistics
modelsummary(
  list("Gaussian(Normal)" = model_developed),
  statistic = "mad",  # Median absolute deviation
  fmt = 2,     # Format results to 2 decimal places
  output = "kableExtra"
) %>%
  kable_styling(font_size = 8.5,
                full_width = FALSE,  # Keeps the table from spanning the full width
    position = "center", # Centers the table
    bootstrap_options = c("striped", "condensed") # Makes rows tighter
  )

@@

With an R-squared value of 0.976, the model demonstrates a strong level of explanatory power, supporting the accuracy of the results. The adjusted R-squared value is 0.975, which is nearly identical to the R-squared value, indicating the model's robustness with 3,312 observations. The RMSE of 4.37 reflects the average magnitude of the model's prediction errors, suggesting good predictive accuracy. Although values like LOOIC and WAIC are excluded due to the lack of alternative models for comparison, the overall results indicate that the model performs well on the cleaned dataset.Further evaluation of the model results will be done in @sec-model-details.

# Discussion {#sec-discussion}

@@

From the summary and analysis of @tbl-sum_sta, we have seen that high income group has the significant high life expectancy compared to the lower income group, which indicates that life expectancy is significantly influenced by income groups. It is known that income determines access to resources, healthcare and living conditions. Difference in life expectancy across income groups reflect broader inequalities in socioeconomic factors and highlight disparities in global health outcomes.

@@

# Appendix {#sec-appx}

@@

## Data Cleaning Table

```{r}
#| label: tbl-cleaned_average
#| tbl-cap: "Cleaned data of average life expectancy of 6 years by different countries. Full cleaned Average life expectancy by country by gender could be find in https://github.com/wendyhuan/lifeexpectancy/blob/main/data/02-analysis_data/average_life_expectancy.csv"
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create a nicely formatted kable table
bothg_average_life_expectancy %>%
  kable(
    col.names = c("Country", "Income Group", "Region", "Average Life Expectancy"),
    align = c("l", "l", "l", "l", "r"),
    digits = 2,
    output = "kableExtra"
) %>%
  kable_styling(font_size = 8.5,
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "condensed")
  ) 
```

@@

**1. Outcome variable (**$y_i$)

@@

**2. Linear Predictor (**$μ_i$)

The linear predictor $μ_i$ is a linear combination of the predictors, which includes an intercept term $α$ and the coefficients $β_1$, $β_2$, $β_3$, corresponding to Region, Income Group, and Gender, respectively. This linear equation models the relationship between life expectancy and the predictors $\mu_i = \alpha + \beta_1(Region) + \beta_2(Income Group) + \beta_3 (Gender)$. The values for these coefficients (betas) represent the expected change in life expectancy for a one-unit change in each respective predictor.

@@

**4. Residual Standard Deviation (**$σ$) The model also estimates $σ$, which represents the standard deviation of the residuals — the errors between the observed life expectancy values and those predicted by the model. The prior for $σ$ is assumed to follow an exponential distribution with a rate of 1, meaning that smaller values of $σ$ (indicating less variation) are somewhat more likely than larger ones.

@@

```{r}
#| label: tbl-testpredicted
#| tbl-cap: "Partical table of the test and training RMSE of Life expectancy over countries and gender of average of 6 years (First 100 rows). Through the model part we could tell that with comparing the training and testing data, the model is well generated without doubt."
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# Format the table using kable
results_df %>%
  head(100) %>%
  kable(col.names = c("Country", "Actual Life Expectancy", "Predicted Life Expectancy", "RMSE (Test)", "RMSE (Train)", "Indication"),
  align = c("l", "r", "r", "r", "r", "l"),
  digits = 2,
  output = "kableExtra"
) |>
  kable_styling(font_size = 7.5,
                full_width = FALSE,  # Keeps the table from spanning the full width
    position = "center", # Centers the table
    bootstrap_options = c("striped", "condensed"))
